### 2025-03-01 cognitive-bias.qmd
20250227 - clean up
20250301 - punting on 4th down

The loss aversion ratio indicates the extent to which a person weighs losses greater than gains, and for many people is around 2 (typically between 1.5 to 2.5; [@Kahneman2011].
That is, people tend to give as much twice the weight to losses as to gains.
Loss aversion is different from [risk aversion](#sec-cognitiveBiasesRiskAversion).
Loss aversion is exemplified when teams play conservatively so as "not to lose" instead of "to win."
In fantasy football, loss aversion may lead managers to start or hold onto players for too long who were highly drafted yet are underperforming instead of starting a more promising player out of fear of losing potential value from their initial investment.
Loss aversion can also influence trade negotiations.

>>>>

The loss aversion ratio indicates the extent to which a person weighs losses greater than gains, and for many people is around 2 (typically between 1.5 to 2.5; [@Kahneman2011].
That is, people tend to give as much twice the weight to losses as to gains.
The late baseball manager Sparky Anderson once said, "Losing hurts twice as bad as winning feels good."
Loss aversion is different from [risk aversion](#sec-cognitiveBiasesRiskAversion).
Loss aversion is exemplified when teams play conservatively so as "not to lose" instead of "to win."
In fantasy football, loss aversion may lead managers to start or hold onto players for too long who were highly drafted yet are underperforming instead of starting a more promising player out of fear of losing potential value from their initial investment.
Loss aversion can also influence trade negotiations.

An example of loss aversion bias in coaching basketball is removing a star player from a game so you can keep the player for the end of the game [@Moskowitz2011].
As described in @sec-coachingGoingForIt, an example of loss aversion bias in football is punting on fourth down when, in many cases, it would be advantageous to go for it [@Moskowitz2011].

### 2025-03-01 sports-cognitive-psychology.qmd
20250301 - punting on 4th down

It had been known for a long time that going for it on fourth down would frequently increase a team's chances of winning.
Despite that, historically, teams rarely went for it on fourth down and elected to punt or kick a field goal instead.
It is curious that there was such a discrepancy between the decisions that would maximize teams' winning percentage and the decisions coaches actually made.
One potential explanation for the discrepancy is because of coaches' risk aversion.
As noted in @sec-cognitiveBiasesRiskAversion, when it is possible to experience either a gain or a loss from a decision, [loss aversion bias](#sec-cognitiveBiasesLossAversion) tends to lead people to make risk-averse decisions [@Kahneman2011].
According to this idea, in the case of failing to successfully convert on fourth down, coaches do not want to have to defend their decision to go for it to the media or the owner or general manager.
That is, they may often play *not to lose*, rather than to win, in order to keep their job.

Rates of going for it on fourth down were low until 2017.
After the 2017 season, rates of going for it on fourth down increased dramatically, as depicted in @fig-4thDownAttemptsByYear.

```{r}
#| label: fig-4thDownAttemptsByYear
#| fig-cap: "The Proportion of Fourth Downs That are Attempts to go for it (Rather Than Punts or Field Goals)."
#| fig-alt: "The Proportion of Fourth Downs That are Attempts to go for it (Rather Than Punts or Field Goals)."

@@

>>>>

### Going for It on Fourth Down {#sec-coachingGoingForIt}

It had been known for a long time that going for it on fourth down would frequently increase a team's chances of winning.
Despite that, historically, teams rarely went for it on fourth down and elected to punt or kick a field goal instead.
It is curious that there was such a discrepancy between the decisions that would maximize teams' winning percentage and the decisions coaches actually made.
In many fourth down situations, by punting the ball, coaches actively and systematically made decisions that reduced their team's chances of winning.
One potential explanation for the discrepancy is because of coaches' risk aversion.
As noted in @sec-cognitiveBiasesRiskAversion, when it is possible to experience either a gain or a loss from a decision, [loss aversion bias](#sec-cognitiveBiasesLossAversion) tends to lead people to make risk-averse decisions [@Kahneman2011].
According to this idea, in the case of failing to successfully convert on fourth down, coaches do not want to have to defend their decision to go for it to the media or the owner or general manager.
That is, they may often play *not to lose*, rather than to win, in order to keep their job.
However, in many cases, this is the wrong choice [@Moskowitz2011].

Rates of going for it on fourth down in the NFL were low until 2017.
After the 2017 season, rates of going for it on fourth down increased dramatically, as depicted in @fig-4thDownAttemptsByYear.

```{r}
#| label: fig-4thDownAttemptsByYear
#| fig-cap: "The Proportion of Fourth Downs That are Attempts to Go for it (Rather Than Punts or Field Goals)."
#| fig-alt: "The Proportion of Fourth Downs That are Attempts to Go for it (Rather Than Punts or Field Goals)."

@@

NFL teams were later to implement analytics into their decision making than even teams at lower levels of the sport.
Some high school teams, such as the Pulaski Academy Bruins (Little Rock, AR), coached by Kevin Kelley, consistently went for it on fourth down well before 2017 [@Moskowitz2011].

> [Bruins Head Football Coach Kevin] Kelley believes that the "quant jocks" don't go far enough to validate the no-punting worldview and, more generally, the virtues of risk-taking.
> "The math guys, the astrophysicist guys, they just do the raw numbers and they don't figure emotion into it—and that's the biggest thing of all," he says.
> "The built-in emotion involved in football is unbelievable, and that's where the benefits really pay off."
> What he means is this: A defense that stops an opponent on third down is usually ecstatic.
> They've done their job.
> The punting unit comes on, and the offense takes over.
> When that defense instead gives up a fourth-down conversion, it has a hugely deflating effect.
> At Pulaski's games, you can see the shoulders of the opposing defensive players slump and their eyes look down when they fail to stop the Bruins on fourth down.
> 
> [@Moskowitz2011, p. 37]

Based on Romer's [-@Romer2006] analysis of third down plays (because few teams went for it on fourth down), focusing on plays in the first quarter (to remove desperation plays), he identified several general conclusions about fourth-down situations:

1. Inside the opponent's 45-yard line, a team is better off going for it than punting with 6 (or less) yards to go (for a first down).
1. Inside the opponent's 33-yard line, a team is better off going for it with 10 (or less) yards to go (unless little time remains and a field goal would decide the game).
1. Once reaching the opponent's 5-yard line, a team is better off going for it.
1. Regardless of field position, a team is always better off going for it with 3 (or less) yards to go.

Nevertheless, out of the fourth down plays @Romer2006 identified between 1998–2000 in which it would have been advantageous to go for it, the team made the suboptimal decision (punting or kicking) 90% of the time.
"Inasmuch as an academic paper can become a cult hit, Romer's made the rounds in NFL executive offices, but most NFL coaches seemed to dismiss his findings as the handiwork of an egghead, polluting art with science." [@Moskowitz2011, p. 39].

Here is an analysis by the New York Times on when to go for it on fourth down: <https://www.nytimes.com/2014/09/05/upshot/4th-down-when-to-go-for-it-and-why.html> (archived at <https://perma.cc/KA9Y-BRUD>).
Here is a fourth down calculator: <https://rbsdm.com/stats/fourth_calculator/>.

::: {#fig-4thDownAnalysis}
![](images/4thDownAnalysis.png){fig-alt="Analysis of When to Go for it on Fourth Down Versus What Coaches Actually Do on 4th Down (From 2014)."}

Analysis of When to Go for it on Fourth Down Versus What Coaches Actually Do on 4th Down (From 2014). Retrieved from <https://www.nytimes.com/2014/09/05/upshot/4th-down-when-to-go-for-it-and-why.html> (archived at <https://perma.cc/KA9Y-BRUD>).
:::

## Other Coaching Absurdities {#sec-coachingAbsurdities}

### Icing the Kicker {#sec-icingTheKicker}
